---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

sapply(c("sim_data_01.rda", "sim_result_01.rda"), function(x) {
  load(file.path(system.file("extdata", package = "rangr"), x),
    envir = globalenv()
  )
})




library(rangr)
```

# rangr

<!-- badges: start -->

[![Status at rOpenSci Software Peer Review](https://badges.ropensci.org/595_status.svg)](https://github.com/ropensci/software-review/issues/595) [![Project Status: Active -- The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active) [![R-CMD-check](https://github.com/ropensci/rangr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ropensci/rangr/actions/workflows/R-CMD-check.yaml) [![codecov](https://codecov.io/gh/ropensci/rangr/graph/badge.svg?token=noq7E0HSe2)](https://app.codecov.io/gh/ropensci/rangr) [![DOI](https://zenodo.org/badge/645848364.svg)](https://zenodo.org/doi/10.5281/zenodo.10569367) 
[![Static Badge](https://img.shields.io/badge/Publication-10.1111%2F2041--210X.14475-94ae63)](https://doi.org/10.1111/2041-210X.14475)


<!-- badges: end -->

The **rangr** package is designed to simulate a species range dynamics. This new tool mimics the essential processes that shape population numbers and spatial distribution: local dynamics and dispersal. Simulations can be run in a spatially explicit and dynamic environment, facilitating population projections in response to climate or land-use changes. By using different sampling schemes and observational error distributions, the structure of the original survey data can be reproduced, or pure random sampling can be mimicked.

The study is supported by the National Science Centre, Poland, grant no. 2018/29/B/NZ8/00066 and the Poznań Supercomputing and Networking Centre (grant no. 403).

# Installation

## Released version

**rangr** is available from CRAN, so you can install it with:

```{r instalation_cran, eval=FALSE}
install.packages("rangr")
```


## Development version

You can install the development version from R-universe or GitHub with:

```{r instalation_dev, eval=FALSE}
install.packages("rangr", repos = "https://ropensci.r-universe.dev")
# or
devtools::install_github("ropensci/rangr")
```

# Basic simulation

Here's an example of how to use the `rangr` package.

## Input maps

Example maps available in rangr in the Cartesian coordinate system:

-   `n1_small.tif`
-   `n1_big.tif`
-   `K_small.tif`
-   `K_small_changing.tif`
-   `K_big.tif`

Example maps available in rangr in the longitude/latitude coordinate system:

-   `n1_small_lon_lat.tif`
-   `n1_big_lon_lat.tif`
-   `K_small_lon_lat.tif`
-   `K_small_changing_lon_lat.tif`
-   `K_big_lon_lat.tif`

```{r SP1.0, eval=FALSE, include=FALSE}
#' @srrstats {SP1.0} Specified domain of applicability
```

You can find additional information about these data sets in help files:

```{r help_input_maps, eval=FALSE}
library(rangr)

?n1_small.tif
?K_small.tif
```

Two of the available datasets, `n1_small.tif` and `K_small.tif`, represent the abundance of a virtual species at the starting point of a simulation and the carrying capacity of the environment, respectively. Both of these objects refer to the same relatively small area, so they are ideal for demonstrating the usage of the package. To view these maps and their dimensions, you can use the following commands:

```{r small_data, warning=FALSE}
library(terra)

n1_small <- rast(system.file("input_maps/n1_small.tif", package = "rangr"))
K_small <-  rast(system.file("input_maps/K_small.tif", package = "rangr"))
```

You can also use the `plot` function from the `terra` package to visualize these maps:

```{r vis_input_maps, fig.align='center', message=FALSE, out.width='70%'}
plot(c(n1_small, K_small))
```

## Initialise

To create a `sim_data` object containing the necessary information to run a simulation, use the `initialise()` function. For example:

```{r init}
sim_data_01 <- initialise(
  n1_map = n1_small,
  K_map = K_small,
  r = log(2),
  rate = 1 / 1e3
)
```

Here, we set the intrinsic population growth rate to `log(2)` and the rate parameter that is related to the kernel function describing dispersal to `1/1e3`.

To see the summary of the `sim_data` object:

```{r summary_sim_data}
summary(sim_data_01)
```

## Simulation

To run a simulation, use the `sim()` function, which takes a `sim_data` object and the specified number of time steps as input parameters. For example:

```{r sim, eval=FALSE}
sim_result_01 <- sim(obj = sim_data_01, time = 100)
```

To see the summary of the `sim_result_01` object:

```{r summary_sim_res, fig.align='center', message=FALSE, out.width='70%'}
summary(sim_result_01)
```

Note that this is a simple example and there are many more parameters that can be set for `initialise()` and `sim()`. See the documentation for the `rangr` package for more information.

## Visualisation

You can use `rangr` to visualise selected time steps from the simulation. The `plot()` method is used to generate the plot. Here's an example:

```{r vis_sim_res_01, warning=FALSE, fig.align='center', message=FALSE, out.width='70%'}
# generate visualisation
plot(sim_result_01,
  time_points = c(1, 10, 25, 50),
  template = sim_data_01$K_map
)
```

You can adjust the `breaks` parameter to get more breaks on the colorscale:

```{r vis_sim_res_02, warning=FALSE, fig.align='center', message=FALSE, out.width='70%'}
# generate visualisation with more breaks
plot(sim_result_01,
  time_points = c(1, 10, 25, 50),
  breaks = seq(0, max(sim_result_01$N_map + 5, na.rm = TRUE), by = 5),
  template = sim_data_01$K_map
)
```

If you prefer working on raster you can also transform any `sim_result` object into `SpatRaster` using `to_rast()` function:

```{r vis_sim_res_03, warning=FALSE, fig.align='center', message=FALSE, out.width='70%'}
# raster construction
my_rast <- to_rast(
  sim_result_01,
  time_points = 1:sim_result_01$simulated_time,
  template = sim_data_01$K_map
)

# print raster
print(my_rast)
```

And then visualise it using `plot()` function:

```{r vis_sim_res_04, warning=FALSE, fig.align='center', message=FALSE, out.width='70%'}
# plot selected time points
plot(my_rast, c(1, 10, 25, 50))
```

# Vignettes

-   [Workflow examples](https://docs.ropensci.org/rangr/articles/rangr.html)

# Citation

To cite `rangr` use `citation()` function:

```{r citation, eval=FALSE}
library(rangr)
citation("rangr")
```

# Code of Conduct

Please note that this package is released with a [Contributor Code of Conduct](https://ropensci.org/code-of-conduct/). By contributing to this project, you agree to abide by its terms.
